
## Introduction

This report examines the relationship between **Attainment 8 Scores** and **House Prices** in **Dorset and Devon**. The analysis involves merging house price data with school attainment data using **PCON_NAME as the linking key**, followed by data cleaning, outlier removal, and linear regression analysis.

---

## Load Required Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load necessary libraries
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(broom)
library(data.table)
```

```{r}
# Load house price dataset
house_prices <- fread("house_price_2022_with_lsoa.csv", 
                      select = c("price", "date", "post", "county", "lsoa11cd"))

# Load school attainment data
attainment_devon <- fread("devon2022.csv")
attainment_dorset <- fread("dorset2022.csv")

# Print column names for debugging
print(names(attainment_devon))
print(names(attainment_dorset))

# Ensure both datasets have the same columns
common_cols <- intersect(names(attainment_devon), names(attainment_dorset))
attainment_devon <- attainment_devon[, ..common_cols]
attainment_dorset <- attainment_dorset[, ..common_cols]

# Combine both datasets safely
attainment_data <- rbind(attainment_devon, attainment_dorset, fill=TRUE)
```

```{r}
# Convert column names to lowercase
setnames(house_prices, tolower(names(house_prices)))
setnames(attainment_data, tolower(names(attainment_data)))

# Standardize LSOA Code column
setnames(attainment_data, "pcon_name", "lsoa11cd")

# Convert LSOA Code to character for accurate merging
house_prices[, lsoa11cd := as.character(lsoa11cd)]
attainment_data[, lsoa11cd := as.character(attainment_data$lsoa11cd)]

# Ensure correct column name for Attainment 8 Score
target_col <- grep("att8", names(attainment_data), value = TRUE, ignore.case = TRUE)

# If multiple columns detected, take the first match
if (length(target_col) > 0) {
    setnames(attainment_data, target_col[1], "att8scr")
} else {
    stop("Column related to 'ATT8SCR' not found in attainment_data. Please check column names.")
}

# Convert Attainment 8 Score to numeric
attainment_data[, att8scr := as.numeric(att8scr)]

# Ensure column exists before aggregation
if ("att8scr" %in% names(attainment_data)) {
    attainment_agg <- attainment_data[, .(attainment8_score = mean(att8scr, na.rm = TRUE)), by = lsoa11cd]
} else {
    stop("Column 'att8scr' still not found. Please verify dataset structure.")
}
```

```{r}
# Merge house prices with attainment data using LSOA code
merged_data <- merge(house_prices, attainment_agg, by = "lsoa11cd", all.x = TRUE)

# Remove rows with missing values
merged_data <- merged_data[!is.na(attainment8_score) & !is.na(price)]

# Convert county to factor
merged_data[, county := as.factor(county)]

# Debugging: Check unique counties
print(unique(merged_data$county))

# Summary of cleaned dataset
summary(merged_data)
```

```{r}
# Remove outliers using the IQR method
Q1 <- quantile(merged_data$price, 0.25)
Q3 <- quantile(merged_data$price, 0.75)
IQR <- Q3 - Q1

# Define bounds for outlier removal
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Filter data within bounds
cleaned_data <- merged_data[price >= lower_bound & price <= upper_bound]

# Debugging: Check data size after outlier removal
print(nrow(cleaned_data))
```

```{r}
# Ensure cleaned data is not empty before plotting
if (nrow(cleaned_data) > 0) {
  ggplot(cleaned_data, aes(x = attainment8_score, y = price, color = county)) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    labs(title = "House Prices vs Attainment 8 Score",
         x = "Attainment 8 Score",
         y = "House Price (£)",
         color = "County") +
    theme_minimal()
} else {
  print("No data available after outlier removal. Adjust filtering criteria.")
}
```

```{r}
# Ensure county has at least two levels before modeling
if (length(unique(cleaned_data$county)) > 1) {
  model <- lm(price ~ attainment8_score + county, data = cleaned_data)
  
  # Model summary
  summary(model)
} else {
  print("County variable has only one level. Ensure both Dorset & Devon data are present.")
}
```

```{r}
# Predict house prices using the model (if model was created)
if (exists("model")) {
  cleaned_data[, predicted_price := predict(model, newdata = cleaned_data)]
  
  # Residual plot
  ggplot(cleaned_data, aes(x = predicted_price, y = price - predicted_price)) +
    geom_point(alpha = 0.6) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    labs(title = "Residuals Plot: House Price Predictions",
         x = "Predicted Price (£)",
         y = "Residual (Actual - Predicted)") +
    theme_minimal()
} else {
  print("Model not created due to insufficient data.")
}
```

```{r}
# Save the cleaned dataset with predictions
if (nrow(cleaned_data) > 0) {
  fwrite(cleaned_data, "house_price_attainment8_model_data.csv")
  cat("Final dataset saved as 'house_price_attainment8_model_data.csv'.")
} else {
  print("No data available to save.")
}
```

---

### **Summary:**
- Cleaned and merged **house price** and **attainment 8 score** datasets.
- Checked for empty datasets before plotting.
- Ensured `county` has at least **two levels** before modeling.
- Removed **outliers** using the **IQR method**.
- Built a **linear regression model** to analyze the relationship.
- Visualized **house prices vs. attainment scores** and **residuals** for model accuracy.

**End of Report.**
